Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-06-07 jss/EpiCompare-paper/EpiCompare-paper.Rmd
major section 4 edits
also cleaned up end of section 3

**Comparisons and assessment**
\shannon{[[ NEW TEXT}

\ben{\sout{Finally,} Often}\footnote{\ben{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}} the data analysis pipeline \ben{\sout{often}} ends with plots, tables, and summary statistics that are used to \sout{\ben{compare} \shannon{and assess}} \ben{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.} In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and into the lens of time-invariant analysis.  We have found \ben{that} these tools \ben{\sout{to be} are} specifically applicable for situations where only one season or cycle of an epidemic has occurred, \ben{or is the object of interest}.

\footnote{\ben{[Ben says: This changed introduction was done to highlight that we can use prediction regions for mutiple types of comparisons.]}}\ben{The first set of of tools surround the creation of prediction regions. We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory. We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic.} [For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations. In this visual setting, comparing this prediction \ben{region} to the true epidemic trajectory \ben{\sout{or comparing the prediction regions defined by two different models' simulations}}\footnote{\ben{[Ben says: this part will be highlighted in later sentences.]}} can be done by eye. In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model. For both of these geometric structures, we provide functions to check if a path is contained (`contained()`)]. \ben{We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are. In \pkg{EpiCompare} we provide the} `hausdorff_dist()` \ben{function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.}

[We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] \ben{We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with} `mean(sim_scores > truth_score)`. [Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.]

@@

[[\textcolor{red}{ OLD TEXT}

One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments. If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space. For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations. In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done by eye. In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model. For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).  The vignettes for `EpiCompare` go into much more mathematical detail about how these functions can be used in conjunction with time-invariant analysis.

We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.

[[\textcolor{red}{Really OLD TEXT}

Finally, in \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC). Corresponding\footnote{\ben{[Ben says: "corresponding to" doesn't seem to be the correct phrase here. If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect word choice.]}\shannon{"To support the ideas presented in Sec."?}} to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations. We have found these tools to be specifically applicable for situations where only one "season" or "cycle"\footnote{\ben{[Ben says: I think we can remove the paratheses here.]} \shannon{accepted}} of the epidemic has occurred.

\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\shannon{motivate better}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments. If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space. For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations. In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done \ben{by eye.\sout{with the 'eye test.'}\shannon{I'm not saying 'no' but I'd do it in an appendix and not the main text.  I don't think equations belong in Section 3.}} In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model. For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).

\footnote{\ben{[Ben says: random thought (not needed) - maybe we do provide some equations for these measures and also maybe mathematically express how we'd examine extremeness?]}}We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations. 
]]

# A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}

@@

\footnote{\ben{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.  Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline \footnote{\ben{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]}} \ben{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\ben{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that? / what does it add to the conv/ why should they reference it?]}}.  Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak\footnote{\ben{[Ben says: I'm not sure this is necessary to state. Additionally - is it really true do we highlight this?]}}.

@@

The next step is to compare the simple SIR model to the EpiModel SIR model. We provide\footnote{\ben{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand. Moreover section 3 already phrases things this way.]}} the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.

>>>>

**Comparisons and assessment**

The last step of the data analysis pipeline often ends with plots, tables, and summary statistics that are used to assess model performance and compare across models or simulations. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and into the lens of time-invariant analysis.  We have found that these tools are specifically applicable for situations where only one season or cycle of an epidemic has occurred or is the object of interest.

The first set of of tools surround the creation of prediction regions. We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory. We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic. For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations. In this visual setting, comparing this prediction region to the true epidemic trajectory can be done by eye. In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model. For both of these geometric structures, we provide functions to check if a path is contained (`contained()`)]. We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are. In \pkg{EpiCompare} we provide the `hausdorff_dist()` function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.

 We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with `mean(sim_scores > truth_score)`. [Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.

# A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}
\shannon{I have substantially overhauled section 4.  The older versions are kept after the entirety of this section but note that much of the text (but not the structure) of this section is different.  I've tried to better motivate why we use the epicompare functions to answer our questions of interest.}

\shannon{[[NEWEST TEXT}
To conclude our paper, we demonstrate how \pkg{EpiCompare}  can streamline the data analysis process with a case study of a measles outbreak in 1861-1862 Germany.  With the help of \pkg{EpiCompare}, we can answer questions such as is a SIR model a good fit for the data, does the outbreak spread differently within the different school classes of the children, and can incorporating an underlying network structure enhance model fit.  We first begin with some context for the epidemic.

## Background for 1861-1862 measles epidemic

 We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.  The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.  In this outbreak, 188 children were infected with measles over the course of three months. This data set includes a rich collection of features including day of measles rash, age, sex, school class, household and household location, and alleged infector of each child.  We show a subset of the data in Table \ref{tab:hags-people}.  We are particularly interested if and how group structure plays a role in the spread of measles.

## Pre-processing and EDA

We begin our analysis by first examining and transforming the raw data (\code{hagelloch_raw}), which is  individual-level data.  With \pkg{EpiCompare}, we can quickly determine how the epidemic progresses through the  population at an aggregate level. By specifying the time of infection (\code{tI}) and the time of recovery (\code{tR}), the function \code{agents_to_aggregate()} calculates the number susceptible, infectious, and recovered individuals at each time step. Once aggregated, we can plot the SIR values through a time-invariant lens using \pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig. \ref{fig:hag-tern-raw}) or with our custom \code{geom}, \code{geom_aggregate()}, which takes the raw agent data as input.

In Figure \ref{fig:hag-tern-raw}, we can focus on the infections over time by analyzing the vertical axis.  Specifically, we see two peaks of infection.  This is interesting because the SIR model, which is sometimes used to model the spread of measles, generally has one defined peak of infection.  We may wonder if the two peaks in the observed data may be due to random noise or if a model more complex than the simple SIR is needed to adequately capture these two peaks.

Previous study tells us that measles outbreaks are often associated with children within the same grade level, and we examine if this is the case here. By combining the function \code{facet_wrap()} with \code{geom_aggregate()} we can easily analyze this scenario,

Immediately in Fig. \ref{fig:tern-class-data}, we see that time invariant infection curve is different for the pre-school class compared to the 1st class.  In the 1st class, we see about 95\% of the class become infected and less than 10\% of them having recovered, which may be indicative of a super-spreading event.  This suspicion is further supported in that `r max(table(hagelloch_raw$IFTO[hagelloch_raw$CL == "1st class"]))` of the `r nrow(hagelloch_raw[hagelloch_raw$CL == "1st class",])` 1st class students have been reportedly infected by the same individual.  We now have some evidence that class structure may play a role in the spread of infection.  We can further analyze this claim with modeling and simulation.

## Modeling and Simulation

We first try to model the Hagelloch data with a baseline stochastic SIR model, which we refer to as the 'simple SIR.' In our full vignette \shannon{(available online),} we show how to fit this simple SIR model via maximum likelihood, a common approach used to fit parameters, and simulate from the model with those best fit parameters. Our function \code{simulate_agents()} (or \code{simulate_SIR_agents()})  generates individual level data according to discrete-time multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.  For example, the below code generates 100 simulations of an outbreak of a disease with one initial infector in a population of $n= 188$ individuals, a scenario analogous to the actual outbreak.

```
```{r}
agg_model <- agents %>% group_by(sim) %>%
  agents_to_aggregate(states = c(I, R)) %>%
  mutate(Type = "Simple SIR")
```
The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` data frame, which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.  Ultimately, we would like to know if this simple SIR model is a good fit to the data, especially in comparison to a SIR model where we can incorporate the class structure.

To fit a more complex SIR model with a network structure, we use package \pkg{EpiModel} \citep{Jenness2018}.  The below code sets up a network of individuals (which includes class as a variable) and then simulates infections over this network.

The output of this network model is \code{epimodel_sir}, an object of class \code{`r class(epimodel_sir)`}, which contains a plethora of modeling information. In the following section, we will use the capabilities of \pkg{EpiCompare} to streamline the process of comparing the two models contained in the objects \code{agg_model} (the simple SIR) and \code{epimodel_sir} (the network SIR model).

The next step is to compare the simple SIR model to the EpiModel SIR model. The \pkg{EpiCompare} function \code{fortify_aggregate()}, takes in an object from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transforms it into a tidy-style data frame.

With the two modeling objects both in the same format, we can then compare the models side-by-side.  The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction region is estimated for the two models.  For the Simple SIR model, we see that while the prediciton region covers the data fairly well, the prediction region clearly misses the second peak of infection.  This indicates that the simple SIR is not a good fit to our data.  We also see that the prediction region is very large, covering up a large area of the ternary plot.  Together, this indicates that the simple SIR model produces a biased model with a large amount of variance.  On the other hand, for the EpiModel network model, we see that the prediction region covers the data quite well and takes up less area.  With \pkg{EpiCompare}, we can see that the model using the class structure is a better fit to the outbreak than the simple SIR model.

The prediction region generated by the network model covers the observed data well, but that does not mean that individual filaments generated from the network model are good fits to the observed data.  We can further examine the model fits by incorporating these filaments into our visualization.  In Fig. \ref{fig:hag-simple-sir} we show the indivdual filaments generated from the two sets of models as gray lines.  Examining these, we see that the individual lines typically only have one defined peak, whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.

We can also examine these filaments quantitatively by using functions that take the distance between the filaments with the observed data.  In the below code, we  transform the simulations to a more computationally-friendly format with the function \code{filament_compression()}. Following that, we calculate the distance between the  simulated filaments and the observed filament with the function \code{dist_matrix_innersq_direction()} and calculate the probability of the truth with respect to those simulations with the function \code{compare_new_to_rest_via_distance()}.  The estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme}).  This indicates that neither of two SIR models are good fits to the data at the filament level.

```{r}
simple_sir <- both_models %>% filter(Type == "Simple SIR") %>%
  rename(S = "X0", I = "X1", R = "X2") %>%
  select(Type, sim, t, S, I, R)

hagelloch_sir2 <- hagelloch_sir %>%
  rename(t = "time") %>%
  mutate(Type = "true observation",
         sim = 0) %>%
  select(Type, sim, t, S, I, R)
```

```{r}
#-- after cleaning up and combining --
all_together_df <- rbind(simple_sir,
                         hagelloch_sir2)
```

```{r hagelloch-extremeness2, echo = FALSE}
simple_sir_info <- simple_sir_true_obs_info %>%
  select(-sim) %>%
  mutate(Type = "Simple SIR")
eimodel_sir_info <- epimodel_sir_true_obs_info %>%
  select(-sim) %>%
  mutate(Type = "EpiModel SIR")

In conclusion, \pkg{EpiCompare} allows us to fully examine this outbreak at every step in the data analysis pipeline (see Fig. \ref{fig:pipeline2}) in a streamlined fashion.  With EDA, we saw evidence that class structure may be important in the spread of of measles.  We then compared a baseline simple SIR model to a more complicated SIR model which incorporated a network structure which included the class structure.  Based on the prediction regions generated from these models, we saw that the network model fit the data much better than the simple SIR model.  However, when we examined the individual filaments generated by the network model, we found that the data is highly unlikely to be generated from such a model.  For further analysis, we would recommend looking into models that can more accurately capture super-spreading events based on the observation that one child was allegedly responsible for nearly all of his classmates' infections.  Overall,  this analysis demonstrates how \pkg{EpiCompare} aids in the data analysis pipeline for both novice and expert practitioners and coders alike.

@@

@@

\footnote{\ben{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.  Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline \footnote{\ben{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]} \shannon{check out new re-write.  new goal is to motivate question of hagelloch and then show how epicompare helps}} \ben{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\ben{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that? / what does it add to the conv/ why should they reference it?]}}.  Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak\footnote{\ben{[Ben says: I'm not sure this is necessary to state. Additionally - is it really true do we highlight this?]}}.

@@

The next step is to compare the simple SIR model to the EpiModel SIR model. We provide\footnote{\ben{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand. Moreover section 3 already phrases things this way.]}\shannon{more of sentence 2 of ben than 1.}} the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.
